---
title: "**Script for data sets import and pre-process**"
author: "João C. D. Muzzi, Juliana F. de Moura, Mauro A. A. Castro, Bonald C. Figueiredo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
geometry: margin=1.5cm
papersize: a4
---

# Data preprocessing
First, we will use TCGABiolinks package to query and download clinical and mRNA-Seq data for TCGA-ACC samples. We will save these information on a SummarizedExperiment R object, where gene names and expression values as well as clinical data for each patient are stored. Then the clinical data will be merged with Sturm et al. (2019) and Thorsson et al. (2018) data sets for TCGA-ACC samples.

# Packages installation and loading
```{r packages-installation}
# Packages Installation
## Find out if packages are already installed, if not proceed with installation
### R packages in CRAN repository
Rpackages <- c("tidyverse","stringr","scales","igraph")
Rpackages.new <- Rpackages[!(Rpackages %in% installed.packages()[,"Package"])]
if(length(Rpackages.new)) install.packages(Rpackages.new)
### R packages in BioConductor
Rpackages <- c("TCGAbiolinks","SummarizedExperiment","AnnotationHub")
Rpackages.new <- Rpackages[!(Rpackages %in% installed.packages()[,"Package"])]
if(length(Rpackages.new)){
  if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
  BiocManager::install(Rpackages.new)
}
```

```{r libraries, message=F, warning=F}
# Libraries used
## For GDC Data Portal ACC RNAseq import and data manipulation
library(TCGAbiolinks)
library(SummarizedExperiment)
## For filtering protein coding genes
library(AnnotationHub)
## Data manipulation
library(tidyverse)
library(stringr)
library(scales)
## Network reconstruction
library(igraph)

```

# GDC Data Portal download with TCGAbiolinks and preprocessing
```{r tcga-download, message=F, warning=F, results=F}
# Download TCGA-ACC cohort data (clinical and RNA-seq) from GDC Portal
query <- GDCquery(project = "TCGA-ACC", 
                  data.category = "Transcriptome Profiling",
                  data.type = "Gene Expression Quantification", 
                  workflow.type = "HTSeq - Counts",
                  sample.type = c("Primary Tumor"))

GDCdownload(query)
# Setting as SummarizedExperiment object 
tcgaACC <- GDCprepare(query, summarizedExperiment = TRUE) # 56457 genes

# Starting to add information to samples
# => Add clinical information to samples
# => Adding TCGA molecular information from marker papers
# => Information will have prefix 'paper_' 
# acc subtype information from:doi:10.1016/j.ccell.2016.04.002
# From the 60483 genes we couldn't map 3881

######################################################### Beginning - Prof Mauro Castro script
##--- Set a 'rowAnnotation' data frame for gene annotation 

ah <- AnnotationHub()
query(ah, "EnsDb")
edb <- query(ah, pattern = c("Homo sapiens", "EnsDb", 101))[[1]]
gns <- genes(edb)
gns$length <- width(gns)
rowAnnotation <- as.data.frame(gns)
rowAnnotation <- rowAnnotation[,c("seqnames","start","end","width",
                                  "strand","gene_id","gene_name","gene_biotype",
                                  "symbol","entrezid","length")] #67990 genes
######################################################## End - Prof Mauro Castro script
```


```{r tcgaACC-mRNA-Seq-Data, results=F, warning=F, message=F}
# Filter 'rowAnnotation' for protein coding genes
rowAnnotation <- dplyr::filter(rowAnnotation, 
                               gene_biotype == "protein_coding")# 22791 genes
# Subset gene expression matrix from SummarizedExperiment object
# for protein coding genes in 'rowAnnotation'
tcgaACC <- subset(tcgaACC, 
                  rownames(tcgaACC) %in% rownames(rowAnnotation)) #19565 genes
# Find duplicated entries
anyDuplicated(rownames(tcgaACC)) # 0

## Select patients with molecular classification on Steroid Phenotype
tcgaACC <- subset(tcgaACC, 
                  select = complete.cases(colData(tcgaACC)$paper_mRNA_K4)) 
# From 79 to 78 patients

rm(ah,edb,gns,query,rowAnnotation)

# Removing zero counts genes
tcgaACC <- tcgaACC[rowSums(assay(tcgaACC))>0,] #19174 genes

# Separating mRNA K4 classification into Steroid Phenotype (High or Low)
# and Proliferation (yes or no)   
colData(tcgaACC)$Steroid <- 
  as.factor(if_else(colData(tcgaACC)$paper_mRNA_K4=="steroid-phenotype-low " |
                      colData(tcgaACC)$paper_mRNA_K4=="steroid-phenotype-low+proliferation",
                    "Steroid_Low",
                    "Steroid_High"))
colData(tcgaACC)$Proliferation_mRNA <- 
  as.factor(if_else(colData(tcgaACC)$paper_mRNA_K4=="steroid-phenotype-high+proliferation" |  
                      colData(tcgaACC)$paper_mRNA_K4=="steroid-phenotype-low+proliferation",
                    "Proliferation_High",
                    "Proliferation_Low"))
colData(tcgaACC)[,c("tumor_stage", "Steroid","Proliferation_mRNA","vital_status")] <-
  lapply(colData(tcgaACC)[,c("tumor_stage", "Steroid","Proliferation_mRNA","vital_status")],
         as.factor)
```

# Download Thorsson and Sturm data sets and additional clinical data
```{r data_prep_Thorsson_Sturm, message=F, warning=F}
# Extract clinical data from SummarizedExperiment object 
clinic.dat <- data.frame(colData(tcgaACC))

# Importing clinical information on hormones excess from:
# "https://api.gdc.cancer.gov/data/a854b659-f0f8-4966-bdc2-279e714983bb/"
repo_link <- "https://api.gdc.cancer.gov/data/"
name.file <- "a854b659-f0f8-4966-bdc2-279e714983bb/"
download.file(
  url = paste0(repo_link,name.file),
  destfile = "clinic_hormone.txt")
clinic <- 
  read.delim(
    file= "clinic_hormone.txt",
    header = T, 
    sep = "\t")
## Filtering for ACC patients barcode
clinic <- clinic[clinic$bcr_patient_barcode %in% clinic.dat$patient,]

## Separating Hormone Excess history in Cortisol and Others
### Retrieve Cortisol excess information for each patient
clinic$Cortisol <- 
  ifelse(grepl("Cortisol",
               clinic$excess_adrenal_hormone_history_type),
         "Cortisol",
         "No")
clinic$Cortisol <-
  ifelse(is.na(clinic$excess_adrenal_hormone_history_type),
         NA,
         clinic$Cortisol)
clinic$Cortisol <- as.factor(clinic$Cortisol)
### Retrieve Other hormones excess for each patient
clinic$Other.Hormones <- 
  ifelse(grepl("Mineral", 
               clinic$excess_adrenal_hormone_history_type),
         "Mineralcorticoids",
         ifelse(grepl(paste0("Androgen","|","Estrogen"),
                      clinic$excess_adrenal_hormone_history_type),
                "Sexual",
                 "No"))
clinic$Other.Hormones <-
  ifelse(is.na(clinic$excess_adrenal_hormone_history_type),
         NA, 
         clinic$Other.Hormones)
clinic$Other.Hormones <- as.factor(clinic$Other.Hormones)

###### Thorsson et al. (2018) data-sets for TCGA samples
# Download Immune Features and Cibersort scores from Thorsson et al 2018
repo_link <- "https://www.cell.com/cms/10.1016/j.immuni.2018.03.023/attachment/"
name.file <- "1b63d4bc-af31-4a23-99bb-7ca23c7b4e0a/mmc2.xlsx"
download.file(
  url= paste0(repo_link,name.file),
  destfile = "thorsson2018.xlsx")
thorsson_pan <- 
  data.frame(readxl::read_xlsx(
    path= "thorsson2018.xlsx",
    col_types = c(rep("text",4),
                  rep("numeric",60))))

## Select patients with Leukocyte Fraction information
thorsson_pan <- thorsson_pan[complete.cases(thorsson_pan$Leukocyte.Fraction),]
## Filter for TCGA-ACC cohort and merge with clinical information
thorsson_ACC <- subset(thorsson_pan,
                   thorsson_pan$TCGA.Participant.Barcode %in% clinic.dat$patient)
thorsson_ACC <- thorsson_ACC[order(thorsson_ACC$TCGA.Participant.Barcode),]
thorsson_ACC$Immune.Subtype <- factor(thorsson_ACC$Immune.Subtype,
                                  levels = c("C1","C2","C3","C4","C5","C6"))

# Download infiltration estimation for TCGA samples (Sturm et al., 2019)
download.file(url = "http://timer.cistrome.org/infiltration_estimation_for_tcga.csv.gz",
              destfile = "sturm2019.csv.gz")
sturm_pan <- 
  read.delim(
    gzfile("sturm2019.csv.gz"),
    header = T, 
    sep = ",")
sturm_pan[,1] <- gsub("-01","",sturm_pan[,1])
names(sturm_pan)[1] <- "patient_XCELL"
## Selecting xCell scores
sturm_pan <- dplyr::select(sturm_pan,
                           contains(c("patient", "XCELL")))
## Filter for TCGA-ACC cohort and merge with clinical information
sturm_ACC <- subset(sturm_pan,
                sturm_pan[,1] %in% clinic.dat$patient)
## Ordering columns of xCell scores by names 
## and separating cells from Immune, Microenv. and Stroma scores
cols <- c(names(sturm_ACC[,2:37])[order(names(sturm_ACC[,2:37]))], names(sturm_ACC[38:40]))
## Merge data-sets for TCGA ACC cohort
tab <- cbind(clinic.dat[order(clinic.dat$patient), "barcode", drop=F],
             clinic[,c("bcr_patient_barcode","Cortisol","Other.Hormones")],
             thorsson_ACC[,1:36],
             thorsson_ACC[,c(37:58)]*thorsson_ACC$Leukocyte.Fraction,
             thorsson_ACC[,59:64],
             sturm_ACC[,c("patient_XCELL", cols)])
tab <- tab[colnames(tcgaACC),]
colData(tcgaACC) <- 
  cbind(colData(tcgaACC)[,c("patient","age_at_index","gender", "tumor_stage",
                            "vital_status","Steroid", "Proliferation_mRNA",
                            "paper_C1A.C1B", "paper_MethyLevel", "paper_Histology")],
        tab)
save(tcgaACC,file = "tcgaACC_pre_processed.RData")
#writexl::write_xlsx(tab, path="/home/muzzi/ACC/ACC_Immune_Scores.xlsx")

```

# Using Extracellular network data from CRI Atlas to reconstruct network
```{r iAtlas-extracellular-network, message=F, fig.align="center", fig.cap="iAtlas Extracellular Communication Network", warning=F, results=F, fig.show="hold"}
##### iAtlas (Thorsson et al 2018) Extracellular communication network reconstruction
##### Importing data from CRI Atlas for TCGA ACC C3 and C4 patients
##### Abundance > 50% and Concordance > 2.5
##### https://isb-cgc.shinyapps.io/shiny-iatlas/ Extracellular Networks

# Network nodes for Abundance > 0.5
download.file(url = "https://github.com/sysbiolab/Sup_Material_Muzzi2021/blob/main/extracellular_network%20_nodes.csv", destfile = "extracellular_network_nodes.csv")
nodes <- 
  read_csv(
    "extracellular_network _nodes.csv", 
    col_types = cols(Abundance = col_number()))
nodes$Type <- as.factor(nodes$Type) # type= cell, ligand or receptor
nodes$Group <- as.factor(nodes$Group) # C3 or C4
## Extracting ligands and receptor from Network for further analysis
nodes_genes <- dplyr::filter(nodes,
                             nodes$Type != "Cell")
nodes_genes <- distinct(nodes_genes, Node, .keep_all = T) 

# Network edges for concordance > 2.5
download.file(url = "https://github.com/sysbiolab/Sup_Material_Muzzi2021/blob/main/extracellular_network_edges.csv", destfile = "extracellular_network_edges.csv")
edges <- 
  read_csv(
    "extracellular_network_edges.csv", 
    col_types = cols(Concordance = col_number()))
edges$Group<- as.factor(edges$Group) # C3 or C4
# Selecting only friendly names
nodes<- nodes[,-1]
edges<- edges[,-c(1,3)]

######### C3 Network
nodesC3 <- nodes[nodes$Group=="C3",]
edgesC3 <- edges[edges$Group=="C3",]

redeC3 <- graph_from_data_frame(d=edgesC3,
                                directed = T,
                                vertices = nodesC3)

########## C4 network
nodesC4 <- nodes[nodes$Group=="C4",]
edgesC4 <- edges[edges$Group=="C4",]

redeC4 <- graph_from_data_frame(d=edgesC4,
                                directed = T,
                                vertices = nodesC4)

# Saving networks for reconstruction in RedeR
networks <- list(redeC3=redeC3,
                 redeC4=redeC4)
save(networks, file= "networks.RData")
```

# Nodes gene expression with Batch corrected and normalized Pan-cancer mRNA-seq data from Hoadley et al. (2018)
```{r heatmap-network-dat-prep, warning=F, results=F}
###############################################################################
### Pan-cancer gene expression of network nodes
###############################################################################
# Heatmap of Network nodes expression
## Retrieve gene names for network nodes
gdc <- c("TCGA", nodes_genes$Node)

# Exporting gene names as tsv file for filtering whole gene exp matrix
write.table(gdc,
            file="/home/muzzi/gdc_genes_filt_symbol.tsv",
            quote = FALSE, 
            row.names = F,
            col.names = F) 
```


```{bash heatmap-network-pancan-exp-filtering}
# Download TCGA Pan cancer, batch corrected and normalized gene expression
# from Hoadley et al. (2018) supplemental data
wget https://api.gdc.cancer.gov/data/9a4679c3-855d-4055-8be9-3577ce10f66e \ 
-o "RNASeqV2-geneExp-xena.tsv"
# Filtering for 25 nodes gene expression in pan-cancer gene expression matrix
grep -Fwf gdc_genes_filt_symbol.tsv 'RNASeqV2-geneExp-xena.tsv' > 'pancan-rna.tsv'
```


```{r pancan-data-prep, warning=F} 
# Data preparation for Network Heatmap
## Importing 24 nodes gene expression matrix for 11,000 TCGA samples
pancan <- read.table(file = "pancan-rna.tsv")
## Rearranging data.frame
pancan <- data.frame(t(pancan))
names(pancan) <- pancan[1,]
pancan <- pancan[-1,]
## Filtering for Primary Tumor samples
pancan <- pancan[grepl("-01$",pancan$sample),]
## Removing duplicated barcodes
pancan <- distinct(pancan, sample, .keep_all = T) #removed 3 duplicated barcodes
## Removing 3 last numbers from Barcodes to merge with Thorsson et al (2018) data set
pancan$sample <- substr(pancan$sample, 1, nchar(pancan$sample)-3)
pancan <- pancan[pancan$sample %in% thorsson_pan$TCGA.Participant.Barcode,] #9361 rows
rownames(pancan) <- pancan[,1]
## Expression values recognized as characters, changing to numberic
pancan[,2:26] <- lapply(pancan[,2:26], 
                        as.numeric)
```

```{r pancan-data, warning=F}
### Saving Pan-cancer preprocessed data
Pancan_data <- list(thorsson_pan=thorsson_pan,
                    sturm_pan=sturm_pan,
                    gen_exp_pan=pancan)
save(Pancan_data, file = "Pancan_data.RData")
```

